---
title: "Method of Simulated Moments with R"
date: "2013-01-29"
categories:
  - R
  - econometrics
toc: true
execute:
  freeze: auto
---

<!-- MathJax scripts -->
<script type="text/javascript" async
  src="https://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>

</head>


<body>

<p>This document details section <em>12.5.6. Unobserved Heterogeneity Example</em>. The original source code giving the results from table 12.3 are available from the authors&#39; site <a href="http://cameron.econ.ucdavis.edu/mmabook/mmaprograms.html">here</a> and written for Stata. This is an attempt to translate the code to R.</p>

<p>Consult the original source code if you want to read the authors&#39; comments. If you want the R source code without all the commentaries, grab it here. This is not guaranteed to work, nor to be correct. It could set your pet on fire and/or eat your first born. Use at your own risk. I may, or may not, expand this example. Corrections, constructive criticism are welcome.</p>

<p>The model is the same as the one described here, so I won&#39;t go into details. The moment condition used is $E[(y_i-\theta-u_i)]=0$, so we can replace the expectation operator by the empirical mean:</p>

<p>$$\dfrac{1}{N} \sum_{i=1}^N(y_i - \theta - E[u_i])=0$$</p>

<p>Supposing that $E[\overline{u}]$ is unknown, we can instead use the method of simulated moments for $\theta$ defined by:</p>

<p>$$\dfrac{1}{N} \sum_{i=1}^N(y_i - \theta - \dfrac{1}{S} \sum_{s=1}^S u_i^s)=0$$</p>

<h3>Import the data</h3>

<p>You can consult the original source code to see how the authors simulated the data. To get the same results, and verify that I didn&#39;t make mistakes I prefer importing their data directly from their website.</p>

```{r}
data <- read.table("http://cameron.econ.ucdavis.edu/mmabook/mma12p2mslmsm.asc")
u <- data[, 1]
e <- data[, 2]
y <- data[, 3]
numobs <- length(u)
simreps <- 10000
```

<h3>Simulation</h3>

<p>In the code below, we simulate the equation defined above:</p>

```{r}
usim <- -log(-log(runif(simreps)))
esim <- rnorm(simreps, 0, 1)

isim <- 0
while (isim < simreps) {

    usim = usim - log(-log(runif(simreps)))
    esim = esim + rnorm(simreps, 0, 1)

    isim = isim + 1

}

usimbar = usim/simreps
esimbar = esim/simreps

theta = y - usimbar - esimbar

theta_msm <- mean(theta)
approx_sterror <- sd(theta)/sqrt(simreps)
```

<p>These steps yield the following results:</p>

```{r}
theta_msm
```

and

```{r}
approx_sterror
```

</body>

